---
title: "Webscraping Feature Films of 2016 from IMDB"
author: "Lucy Murray"
date: "7/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This lab follows [Beginner’s Guide on Web Scraping in R](https://www.analyticsvidhya.com/blog/2017/03/beginners-guide-on-web-scraping-in-r-using-rvest-with-hands-on-knowledge/) using rvest with hands-on example.

"Webscraping is a technique for converting the data present in unstructured format (HTML tags) over the web to the structured format which can easily be accessed and used." For this assignment, we will be examining the [most popular feature films of 2016](https://www.imdb.com/search/title/?count=100&release_date=2016,2016&title_type=feature) from the IMDB website utilizing the "rvest" package and the SelectorGadget Chrome extension.

### Scraping a webpage using R

In order to begin the project, we need to read the html of the web page we are interested in, so that we can begin scraping the data we would like to scrape from the website.

```{r}
#Accessing the rvest library
library("rvest")

#Specifying the url for desired website to be scraped
url <- 'http://www.imdb.com/search/title?count=100&release_date=2016,2016&title_type=feature'

#Reading the HTML code from the website
webpage <- read_html(url)
```

Here is a list of data we are interested in:
* Rank
* Title
* Description
* Runtime
* Genre
* Rating
* Votes
* Director
* Actor
* Metascore
* Gross_Earning_in_Mil

### Scraping the Rankings (Steps 1-4)

After using the CSS SelectorGadget to select only the rankings, I will use the following code to get all of the rankings and save them into the variable rank_data.

```{r}
#Using CSS selectors to scrape the rankings section
rank_data_html <- html_nodes(webpage,'.text-primary')

#Converting the ranking data to text
rank_data <- html_text(rank_data_html)

#Let's have a look at the rankings
head(rank_data)
```

I can use the following code to convert the data into a numerical format, thus making it easier to work with.
```{r}
#Data-Preprocessing: Converting rankings to numerical
rank_data<-as.numeric(rank_data)

#Let's have another look at the rankings
head(rank_data)
```

### Scraping the Titles (Steps 5-6)

After using the CSS SelectorGadget to select only the titles, I will use the following code to get all of the titles and save them into the variable title_data. 
```{r}
#Using CSS selectors to scrape the title section
title_data_html <- html_nodes(webpage,'.lister-item-header a')

#Converting the title data to text
title_data <- html_text(title_data_html)

#Let's have a look at the title
head(title_data)
```

### Scraping the Rest of the Data (Step 7)

I will now repeat the previous steps using the CSS SelectorGadget one at a time to scrape the Description, Runtime, Genre, Rating, Metascore, Votes, Gross_Earning_in_Mil, Director and Actor data.

##### Scraping the Description
```{r}
#Using CSS selectors to scrape the description section
description_data_html <- html_nodes(webpage,'.ratings-bar+ .text-muted')

#Converting the description data to text
description_data <- html_text(description_data_html)

#Let's have a look at the description data
head(description_data)
```

For the description_data, we will clean it by removing the \n before each of the strings.
```{r}
#Data-Preprocessing: removing '\n'
description_data<-gsub("\n","",description_data)

#Let's have another look at the description data 
head(description_data)
```

##### Scraping the Runtime
```{r}
#Using CSS selectors to scrape the Movie runtime section
runtime_data_html <- html_nodes(webpage,'.text-muted .runtime')

#Converting the runtime data to text
runtime_data <- html_text(runtime_data_html)

#Let's have a look at the runtime
head(runtime_data)
```

For the runtime_data, we will clean it by removing the mins string after each value, and by connverting it into a numeric data type.
```{r}
#Data-Preprocessing: removing mins and converting it to numerical
runtime_data<-gsub(" min","",runtime_data)
runtime_data<-as.numeric(runtime_data)

#Let's have another look at the runtime data
head(runtime_data)
```

##### Scraping the Genre
```{r}
#Using CSS selectors to scrape the Movie genre section
genre_data_html <- html_nodes(webpage,'.genre')

#Converting the genre data to text
genre_data <- html_text(genre_data_html)

#Let's have a look at the runtime
head(genre_data)
```

For the genre_data, we will clean it by removing the \n string, removing excess space, taking only the first genre of each movie before each value, and by converting each genre from text to a factor.
```{r}
#Data-Preprocessing: removing \n
genre_data<-gsub("\n","",genre_data)

#Data-Preprocessing: removing excess spaces
genre_data<-gsub(" ","",genre_data)

#taking only the first genre of each movie
genre_data<-gsub(",.*","",genre_data)

#Converting each genre from text to factor
genre_data<-as.factor(genre_data)

#Let's have another look at the genre data
head(genre_data)
```

##### Scraping the IMDB Rating
```{r}
#Using CSS selectors to scrape the IMDB rating section
rating_data_html <- html_nodes(webpage,'.ratings-imdb-rating strong')

#Converting the ratings data to text
rating_data <- html_text(rating_data_html)

#Let's have a look at the ratings
head(rating_data)
```

For the rating_data, we will clean it by converting the values to numeric data types.
```{r}
#Data-Preprocessing: converting ratings to numerical
rating_data<-as.numeric(rating_data)

#Let's have another look at the ratings data
head(rating_data)
```

##### Scraping the Votes Section
```{r}
#Using CSS selectors to scrape the votes section
votes_data_html <- html_nodes(webpage,'.sort-num_votes-visible span:nth-child(2)')

#Converting the votes data to text
votes_data <- html_text(votes_data_html)

#Let's have a look at the votes data
head(votes_data)
```

For the votes_data, we will clean it by removing the commas and converting votes to a numerical data type.
```{r}
#Data-Preprocessing: removing commas
votes_data<-gsub(",","",votes_data)

#Data-Preprocessing: converting votes to numerical
votes_data<-as.numeric(votes_data)

#Let's have another look at the votes data
head(votes_data)
```

##### Scraping the Directors Section
```{r}
#Using CSS selectors to scrape the directors section
directors_data_html <- html_nodes(webpage,'.text-muted+ p a:nth-child(1)')

#Converting the directors data to text
directors_data <- html_text(directors_data_html)

#Let's have a look at the directors data
head(directors_data)
```

For the directors_data, we will clean it by converting it to a factor data type.
```{r}
#Data-Preprocessing: converting directors data into factors
directors_data<-as.factor(directors_data)
```

##### Scraping the Actors Section
```{r}
#Using CSS selectors to scrape the actors section
actors_data_html <- html_nodes(webpage,'.lister-item-content .ghost+ a')

#Converting the gross actors data to text
actors_data <- html_text(actors_data_html)

#Let's have a look at the actors data
head(actors_data)
```

For the actors_data, we will clean it by converting it to a factor data type.
```{r}
#Data-Preprocessing: converting actors data into factors
actors_data<-as.factor(actors_data)
```

##### Scraping the Metascore Section
```{r}
#Using CSS selectors to scrape the metascore section
metascore_data_html <- html_nodes(webpage,'.metascore')

#Converting the runtime data to text
metascore_data <- html_text(metascore_data_html)

#Let's have a look at the metascore 
head(metascore_data)
```

For the metascore_data, we will clean it by removing the extra space in metascore and by checking the length of the the metascore data.
```{r}
#Data-Preprocessing: removing extra space in metascore
metascore_data<-gsub(" ","",metascore_data)

#Lets check the length of metascore data
length(metascore_data)

```

### Inserting NAs for missing Metascore Data (Steps 8-9)

The length of 97 indicates that 3 films are missing a metascore data. By manually searching the page, we see that The Inivisble Guest (#38), Dangal (#61), and Terrifier (#71) are the titles with missing metascore data. We can insert an NA into these rows so that we can still obtain statistic summary data.

```{r}
for (i in c(38,61,71)){

a<-metascore_data[1:(i-1)]
b<-metascore_data[i:length(metascore_data)]
metascore_data<-append(a,list("NA"))
metascore_data<-append(metascore_data,b)

}

#Data-Preprocessing: converting metascore to numerical
metascore_data<-as.numeric(metascore_data)

#Let's have another look at length of the metascore data
length(metascore_data)

#Adding NAs to account for missing data
metascore_data <- c(metascore_data, NA, NA)

#Let's look at summary statistics
summary(metascore_data)
```

### Repeating the Previous Steps for Gross variable (Step 10)

Similar to what happened before, the gross variable representing gross earnings in millions has missing gross data. We can fix this using the same solution as before, by manually finidng where data is missing and inserting NA values.

```{r}
#Using CSS selectors to scrape the gross revenue section
gross_data_html <- html_nodes(webpage,'.ghost~ .text-muted+ span')

#Converting the gross revenue data to text
gross_data <- html_text(gross_data_html)

#Let's have a look at the votes data
head(gross_data)

#Data-Preprocessing: removing '$' and 'M' signs
gross_data<-gsub("M","",gross_data)

gross_data<-substring(gross_data,2,6)

#Let's check the length of gross data
length(gross_data)
```

##### Inserting the NAs for missing Gross data

The length of 92 indicates that 8 films are missing gross data. By manually searching the page, we see that The Invisible Ghost (#38), The Autospy of Jane Doe (#47), The Do-Over (#63), Below Her Mouth (#65), Hush (#69), Terrifier (#71), Mean Dreams (#87), and Brimstone (#94) are the titles with missing gross data. We can insert an NA into these rows so that we can still obtain statistic summary data. We will also convert the data into a numeric type so we can observe the summary statistics.

```{r}
#Filling missing entries with NA
for (i in c(38,47,63,65,69,71,87,94)){

a<-gross_data[1:(i-1)]
b<-gross_data[i:length(gross_data)]
gross_data<-append(a,list("NA"))
gross_data<-append(gross_data,b)

}

#Data-Preprocessing: converting gross to numerical
gross_data<-as.numeric(gross_data)

#Let's have another look at the length of gross data
length(gross_data)
summary(gross_data)
```

### Creating a Dataframe (Step 11)

Now that all of the 11 variables we are interested in have been scraped, we will construct a data frame and inspect its structure.

```{r}
#Combining all the lists to form a data frame
movies_df <- data.frame(
  Rank = rank_data, Title = title_data,
  Description = description_data,
  Runtime = runtime_data,
  Genre = genre_data,
  Rating = rating_data,
  Metascore = metascore_data,
  Votes = votes_data,
  Gross_Earning_in_Mil = gross_data,
  Director = directors_data,
  Actor = actors_data)

#Structure of the data frame
str(movies_df)
```

### Analyzing Scraped Data from the Web

##### Histogram for Runtime of Films by Genre
```{r}
#Initial Histogram
library('ggplot2')
qplot(data = movies_df,Runtime,fill = Genre,bins = 30)

#Histogram with alternative x-axis intervals and labels
qplot(data = movies_df,Runtime,fill = Genre,bins = 30) +
scale_x_continuous(breaks = seq(0, 200, by = 25)) +
  labs(title = "Number of Films vs Runtime by Genre", 
       x = "Runtime (mins)", y = "Number of Films")
```

**Question 1:** Based on the above data, which movie from which Genre had the longest runtime?
**Answer:** The graph shows that the genre with the highest runtime is "Adventure", but to find out the specific film we need to arrange the data by runtime from longest to shortest.

```{r}
#arranging the data by runtime
library("dplyr")
head(arrange(movies_df, desc(runtime_data)))
```
This shows that the movie American Honey from the genre "Adventure" has the longest runtime.

##### Scatterplot for Runtime by Vote Ratings and Genre
```{r}
##Initial Scatterplot
ggplot(movies_df,aes(x=Runtime,y=Rating))+
geom_point(aes(size=Votes,col=Genre))

#Scatterplot with alternative x-axis intervals and labels
ggplot(movies_df,aes(x=Runtime,y=Rating))+
geom_point(aes(size=Votes,col=Genre))+
scale_x_continuous(breaks = seq(0, 200, by = 25)) +
  labs(title = "Vote Rating vs Runtime by Genre and Number of Votes", 
       x = "Runtime (mins)", y = "Vote Rating")
```

**Question 2:** Based on the above data, in the Runtime of 130-160 mins, which genre has the highest votes?
**Answer:** In the runtime of 130-160 mins, the genre "Action" has the highest votes. You can see this by observing that action has the largest circles, which represent number of votes.

##### Scatterplot for Runtime by Vote Ratings, Genre, and Gross Earnings in Millions
```{r}
#Initial Scatterplot
ggplot(movies_df,aes(x=Runtime,y=Gross_Earning_in_Mil))+
geom_point(aes(size=Rating,col=Genre))

#Scatterplot with alternative x-axis intervals and labels
ggplot(movies_df,aes(x=Runtime,y=Gross_Earning_in_Mil))+
geom_point(aes(size=Rating,col=Genre))+
scale_x_continuous(breaks = seq(0, 200, by = 20)) +
  labs(title = "Gross Earnings vs Runtime by Genre and Rating", 
       x = "Runtime (mins)", y = "Gross Earnings (Millions)")
```

**Question 3:** Based on the above data, across all genres which genre has the highest average gross earnings in runtime 100 to 120.
**Answer:** The "Adventure" genre has the highest average gross earnings in runtime 100 to 120.

